### 2025-02-03 paper/preprint/paper.tex
progress baby

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
van Deursen}\\Faculty of Electrical Engineering, Mathematics and
Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations (CE) have emerged as a popular tool to
explain predictions made by opaque machine learning models: they explain
how factual inputs need to change in order for some fitted model to
produce some desired output. Much existing research has focused on
identifying explanations that are not only valid but also deemed
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning desirable explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and desirable explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver desirable
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, I. J. Goodfellow, Shlens, and Szegedy
(\citeproc{ref-goodfellow2014explaining}{2014}) showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AE):
counterfactuals of model inputs that yield vastly different model
predictions despite being semantically indifferent from their factual
counterparts. Despite partially effective mitigation strategies such as
\textbf{adversarial training}, truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).

@@

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems involving opaque
models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
tools can aide us in monitoring models and providing recourse to
affected individuals (\citeproc{ref-wachter2017counterfactual}{Wachter,
Mittelstadt, and Russell 2017}).

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are deemed desirable by humans. Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both desirable and valid for any given
model, independent of whether the learned representations are also
compatible with undesirable explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

@@

\subsection{Background on Counterfactual
Explanations}\label{background-on-counterfactual-explanations}

(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
Russell 2017}; \citeproc{ref-joshi2019realistic}{Joshi et al. 2019};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024})

\subsection{Learning Representations}\label{learning-representations}

\begin{quote}
For example, joint-energy models
\end{quote}

\subsection{Generalization and
Robustness}\label{generalization-and-robustness}

Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
counterfactual images for MNIST and ImageNet through independent
mechanisms (IM): each IM learns class-conditional input distributions
over a specific lower-dimensional, semantically meaningful factor, such
as \emph{texture}, \emph{shape} and \emph{background}. They demonstrate
that using these generated counterfactuals during classifier training
improves model robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization.

Teney, Abbasnedjad, and Hengel (\citeproc{ref-teney2020learning}{2020})
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on, firstly, identifying similar input samples with different
annotations and, secondly, ensuring that the gradient of the classifier
aligns with the vector between pairs of counterfactual inputs using the
cosine distance as a loss function (referred to as \emph{gradient
supervision}) (\textbf{\emph{this might be useful for our task as
well}}). In the natural language processing (NLP) domain,
counterfactuals have similarly been used to improve models through data
augmentation: Wu et al. (\citeproc{ref-wu2021polyjuice}{2021}), propose
POLYJUICE, a general-purpose counterfactual generator for language
models. They demonstrate empirically that augmenting training data
through POLYJUICE counterfactuals improves robustness in a number of NLP
tasks.

\subsection{Link to Adversarial
Training}\label{link-to-adversarial-training}

From this perspective, adversarial training induces models to
``unlearn'' representations that are susceptible to the semantically
most meaningless explanations---adversarial examples.

Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) propose two
definitional differences between Adversarial Examples (AE) and
Counterfactual Explanations (CE): firstly, and more importantly
according to the authors, the term AE implies missclassification, which
is not the case for CE (\textbf{\emph{this might be a useful notion for
use to distinguish between adversarials and explanations during
training}}); secondly, they argue that closeness plays a more critical
role in the context of CE but confess that even counterfactuals that are
not close might be relevant explanations. Pawelczyk et al.
(\citeproc{ref-pawelczyk2022exploring}{2022}) show that CE and AE are
equivalent under certain conditions and derive upper bounds on the
distances between them.

\subsection{Closely Related}\label{closely-related}

Guo, Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the
first to propose end-to-end training pipeline that includes
counterfactual explanations as part of the training prodeduce. In
particular, they propose a specific network architecture that includes a
predictor and CE generator network (\textbf{\emph{akin a GAN?}}), where
the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network (\textbf{\emph{here we are aligned}}). In
contrast, we impose no restrictions on the neural network architecture
at all. (\textbf{\emph{to ensure the one-hot encoding of categorical
features is maintained, they simple use softmax (might be interesting
for CE.jl)}}) Interestingly, the authors find that their approach is
sensitive to the choice of the loss function: only MSE seems to lead to
good performance. They also demonstrate theoretically, that the
objective function is difficult to optimize due to divergent gradients
and suffers from poor adversarial robustness. (\textbf{\emph{because
partial gradients with respect to the classification loss component and
the counterfactual validity component point in opposite directions}}).
To mitigate these issues, the authors use block-wise gradient descent:
they first update with respect to classification loss and then use a
second update with respect to the other loss components
(\textbf{\emph{this might be useful for our task as well}}). Ross,
Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024}) propose a
way to train models that are guaranteed to provide recourse for
individuals with high probability. The approach builds on adversarial
training (\textbf{\emph{here we are aligned}}), where in this context
adversarial examples are actively encouraged to exist, but only target
attacks with respect to the positive class. The proposed method allows
for imposing a set of actionable recourse ex-ante: for example, users
can impose mutability constraints for features (\textbf{\emph{here we
are aligned}}). (\textbf{\emph{To solve their objective function more
efficiently, they use a first-order Taylor approximation to approximate
the recourse loss component (might be applicable in our case)}})

Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
Counterfactual Adversarial Training (CAT) with intention of improving
generalization and robustness of language models. Specifically, they
propose to proceed as follows: firstly, identify training samples that
are subject to high predictive uncertainty (entropy); secondly, generate
counterfactual explanations for those samples; and, finally, finetune
the model on the augmented dataset that includes the generated
counterfactuals.

@@

@@

@@

\bibitem[\citeproctext]{ref-joshi2019realistic}
Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
Actionable Explanations in Black-Box Decision Making Systems.''}
\url{https://arxiv.org/abs/1907.09615}.

@@

@@

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
van Deursen}\\Faculty of Electrical Engineering, Mathematics and
Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed desirable with
respect to the underlying data and stakeholder requirements. Recent work
has shown that under this premise, the task of learning desirable
explanations is effectively reassigned from the model itself to the
(post-hoc) counterfactual explainer. Building on that work, we propose a
novel model objective that leverages counterfactuals during the training
phase (ad-hoc) in order to minimize the divergence between learned
representations and desirable explanations. Through extensive
experiments, we demonstrate that our proposed methodology facilitates
training models that inherently deliver desirable explanations while
maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
showed that artificial neural networks (ANN) are sensitive to
adversarial examples: counterfactuals of model inputs that yield vastly
different model predictions despite being ``imperceptible'' in that they
are semantically indifferent from their factual counterparts. Despite
partially effective mitigation strategies such as \textbf{adversarial
training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).

@@

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems (ADMS) involving
opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
XAI tools can aide us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g.~``loan application
rejected'') into positive ones (``application accepted''). Wachter,
Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) propose
\textbf{counterfactual explanations} as an effective approach to achieve
this: they explain how factual inputs need to change in order for some
fitted model to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed desirable by humans.
Instead, the choice has typically been to improve the capacity of XAI
tools to identify the subset explanations that are both desirable and
valid for any given model, independent of whether the learned
representations are also compatible with undesirable explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed desirable with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some undesirable explanations from the solution
space.

\subsection{Adversarial Examples are Counterfactual
Explanations}\label{adversarial-examples-are-counterfactual-explanations}

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization. Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function.
In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation: Wu et
al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{POLYJUICE}, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through
\emph{POLYJUICE} counterfactuals improves robustness in a number of NLP
tasks. Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023})
introduce Counterfactual Adversarial Training (CAT), which also aims at
improving generalization and robustness of language models.
Specifically, they propose to proceed as follows: firstly, they identify
training samples that are subject to high predictive uncertainty;
secondly, they generate counterfactual explanations for those samples;
and, finally, they fine-tune the given language model on the augmented
dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations (CE) and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
case for jointly studying the opaqueness and robustness problem in
representation learning. Formally, AE can be seen as the subset of CE,
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
CE and AE are equivalent under certain conditions and derive theoretical
upper bounds on the distances between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
propose a way to train models that are guaranteed to provide recourse
for individuals to move from an adverse outcome to some positive target
class with high probability. The approach proposed by Ross, Lakkaraju,
and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
adversarial training, where in this context susceptibility to targeted
adversarial examples for the positive class is explicitly induced. The
proposed method allows for imposing a set of actionability constraints
ex-ante: for example, users can specify that certain features
(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
to propose an end-to-end training pipeline that includes counterfactual
explanations as part of the training procedure. In particular, they
propose a specific network architecture that includes a predictor and CE
generator network, where the parameters of the CE generator network are
learnable. Counterfactuals are generated during each training iteration
and fed back to the predictor network. In contrast to Guo, Nguyen, and
Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
restrictions on the neural network architecture at all.

\subsection{Beyond Robustness}\label{beyond-robustness}

Improving the adversarial robustness of models is not the only path
towards aligning representations with desirable explanations. In a work
closely related to this one, Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
be improved through model averaging and refined model objectives. The
authors propose a way to generate counterfactuals that are maximally
\textbf{faithful} to the model in that they are consistent with what the
model has learned about the underlying data. Formally, they rely on
tools from energy-based modelling to minimize the divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by the model. Their proposed counterfactual explainer,
\emph{ECCCo}, yields desirable (or \textbf{plausible}) explanations if
and only if the underlying model has learned representations that align
with them. They find that both deep ensembles
(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
and Blundell 2017}) and joint energy-based models (JEMs)
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data
(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
mitigates the aforementioned risk of relying on a single locally optimal
representations that corresponds to semantically meaningless
explanations for the data. Previous work by Schut et al.
(\citeproc{ref-schut2021generating}{2021}) similarly found that
generating desirable (``interpretable'') counterfactual explanations is
almost trivial for deep ensembles that have also undergone adversarial
training. The case for JEMs is even clearer: they involve a hybrid
objective that induces both high predictive performance and generative
capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
is closely related to the idea of aligning models with desirable
explanations and has inspired our proposed counterfactual training
objective, as we explain in Section~\ref{sec-method}.

@@

\begin{definition}[Model
Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}

\end{definition}

@@

\bibitem[\citeproctext]{ref-grathwohl2020your}
Grathwohl, Will, Kuan-Chieh Wang, Joern-Henrik Jacobsen, David Duvenaud,
Mohammad Norouzi, and Kevin Swersky. 2020. {``Your Classifier Is
Secretly an Energy Based Model and You Should Treat It Like One.''} In
\emph{International Conference on Learning Representations}.

@@

@@

\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
Ensembles.''} \emph{Advances in Neural Information Processing Systems}
30.

@@

\bibitem[\citeproctext]{ref-szegedy2013intriguing}
Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2013. {``Intriguing
Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.

### 2025-02-03 paper/sections/abstract.qmd
progress baby

Counterfactual Explanations (CE) have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output. Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements. Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer. Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations. Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.

>>>>

Counterfactual Explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output. Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements. Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer. Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations. Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.

### 2025-02-03 paper/sections/intro.qmd
progress baby

This trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts. Despite partially effective mitigation strategies such as **adversarial training**, truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

@@

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons]. Effective XAI tools can aide us in monitoring models and providing recourse to affected individuals [@wachter2017counterfactual].

>>>>

This trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Despite partially effective mitigation strategies such as **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

@@

The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations. This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions. On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems (ADMS) involving opaque models [@oneil2016weapons]. Effective XAI tools can aide us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g. "loan application rejected") into positive ones ("application accepted"). @wachter2017counterfactual propose **counterfactual explanations** as an effective approach to achieve this: they explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.

### 2025-02-03 paper/sections/method.qmd
progress baby

# Related Literature {#sec-lit}

To the best of our knowledge, our proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed desirable with respect to the underlying data and stakeholder requirements. To make this more concrete, we follow @augustin2020adversarial in tieing the concept of explainability to the quality of counterfactual explanations that we can generate for a given model. The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples. This finding is intuitive when we look at adversarial training (AT) as a means of regularization in the context of representation learning with high degrees of freedom: it reduces the space of viable solutions by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e. adversarial examples). @szegedy2013intriguing

The link between counterfactual explanations and adversarial examples has been studied elsewhere: @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.

@@

@altmeyer2024faithful find that

## Background on Counterfactual Explanations

[@wachter2017counterfactual;@joshi2019realistic;@altmeyer2024faithful]

## Learning Representations

> For example, joint-energy models

## Generalization and Robustness

@sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*. They demonstrate that using these generated counterfactuals during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.

@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*) (***this might be useful for our task as well***). In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose POLYJUICE, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.

## Link to Adversarial Training

@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (***this might be a useful notion for use to distinguish between adversarials and explanations during training***); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.

## Closely Related

@guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce. In particular, they propose a specific network architecture that includes a predictor and CE generator network (***akin a GAN?***), where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network (***here we are aligned***). In contrast, we impose no restrictions on the neural network architecture at all. (***to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)***) Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance. They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness. (***because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions***). To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components (***this might be useful for our task as well***). @ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability. The approach builds on adversarial training (***here we are aligned***), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class. The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features (***here we are aligned***). (***To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case)***)

@luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) with intention of improving generalization and robustness of language models. Specifically, they propose to proceed as follows: firstly, identify training samples that are subject to high predictive uncertainty (entropy); secondly, generate counterfactual explanations for those samples; and, finally, finetune the model on the augmented dataset that includes the generated counterfactuals.

>>>>

# Counterfactual Training {#sec-method}

@@
